Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden

### 2024-02-07 analysis/paper/paper.qmd
Small edit
Review corrections

---
title: "Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden"
author:
  - Mattias Sjölander:
      email: mattias.sjolander@umu.se
      institute: [UmU]
      correspondence: true
  - Johan Linderholm:
      email: johan.linderholm@umu.se
      institute: [UmU]
      correspondence: false
  - Paul Geladi:
      email: paul.geladi@slu.se
      institute: [SLU]
      correspondence: false
  - Philip I. Buckland:
      email: philip.buckland@umu.se
      institute: [UmU]
      correspondence: false
institute:
  - UmU: Environmental Archaeology Lab, Department of Historical, Philosophical and Religious studies, Umeå University, Umeå, Sweden
  - SLU: Department of Forest Biomaterials and Technology, Swedish University of Agricultural Sciences, Umeå, Sweden 
title-block-published: "Last updated"  
date: now
date-format: long
crossref: 
  fig-title: '**Fig.**'
  fig-labels: '**arabic**'
  fig-prefix: Fig.
format:
  pdf:
    documentclass: scrartcl
    fig-pos: "H"
    geometry: margin = 2.54cm 
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 300
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/quaternary-science-reviews.csl" # Insert path for the bib-style
abstract: |
  Northern Fennoscandia represents a geologically complex region affected by both glacial and postglacial processes. Quartzite was a key material type utilized by hunter-gatherers in Northern Sweden around the period 4 000 – 2 000 BP, and is thus critical to the understanding of raw material procurement and material flow within the region. However, there is a severe lack of methodological development in the characterization of these materials, and provenance of locally available geological material is complex and fraught with uncertainty. 126 points and preforms made from quartz/quartzite were sampled from 47 archaeological sites along the upper Ångerman river valley in Västerbotten, Sweden. The material has been analysed non-destructively using three separate portable spectroscopic instrumentations (Near-infrared, Raman, X-Ray Fluorescence). Evaluation of the spectra and exploratory data analysis using Principal Component Analysis demonstrates detectable differences in the material that likely stem from diagenetic/paragenetic origin. The presence of graphite, muscovite and biotite could likewise provide information on the material’s metamorphic grade. In addition to reaffirming the potential of field-based screening instrumentation, these results will benefit future surveys of geological sources in the region. They also indicate potential for the construction of a predictive model that could classify the quartzite based on its chemical characteristic. Such a model would prove useful in future spatial analysis and testing of models of raw material management.
keywords: |
  Hunter-Gatherer; Lithics; Provenance; Quartzite; Vibrational spectroscopy; Scandinavia
---

@@

Petrographic characterization and provenance studies are instrumental for inferring trade and exchange among prehistoric societies [@tykot03]. The presence of lithic material foreign to the local region of an archaeological site, raises important questions on the processes that resulted in its eventual deposition and discovery. The characterization of geological material and identification of geological sources are the first steps in reconstructing a prehistoric community’s strategies for raw material selection and management. Flint and obsidian have been central to studies of trade and exchange in European prehistory, due in part to the well-surveyed geological sources for these materials [@tykot17]. This focus has led to a bias which has been highlighted by Prieto et al. [-@prieto_etal19: 15]. Recent decades, however, have seen an increase in interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19].

In Northern Sweden, quartz and quartzite dominate the lithic assemblages and have been used in settlement, exchange and mobility studies [@baudou78]. Models have been proposed for raw material sourcing and trade and exchange for the less common slate and flint finds within Northern Sweden [@halen94: 115-122; @lundberg97: 161-171]. There has, however, been little development in the petrographic characterization and provenance study of quartz and quartzite [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10]. ). A possible reason for this might well be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16]. Northern Sweden is extensively covered by glacial till, and whilst there are identifiable source areas (e.g. small bedrock quarries), there is an extremely large number of other potential sources of raw materials to be found in moraine deposits. The few existing theories on lithic material sourcing within the North Swedish region describe strategies for direct procurement from the mountain zone [@forsberg85; -@forsberg12]. This does not, however, account for the possibility that the material was sourced locally around the site, as part of an embedded adaptation strategy [@binford79; @rankama_etal06: 249-250]. A new approach to provenance studies is therefore required in this region.

@@

![Geographical distribution of the sampled sites, with find context, along the river of Ångermanälven, Västerbotten, Sweden. The Late Quaternary reincorporation of geological material from the Scandian mountains into the sediments in the forest inlands adds to the complexity of questions of provenance. Tectonic symbology after (Gee and Stephens 2020). Made with © Sveriges Geologiska Undersökning, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet.](../figures/Fig1.jpg){#fig-distribution-map width=100%, height=100%}

@@

This study includes `r length(quartzite)` and `r length(bquartz)` *brecciated quartz* samples. The inclusion of *brecciated quartz* in the study enables the separation of quartz from quartzite, based on non-destructive analysis of their chemical and structural characteristics, to be evaluated. The term  *brecciated quartz* (also referred to as *mylonite quartz*) was introduced to archaeologists in Northern Sweden by the geologist Åhman [-@ahman67: 8] in his analysis of the petrographic material in the *Early Norrland* project [@biornstad68; @ahman67]. Åhman describes the material as easily mistaken for an unusually pure quartzite, but where the fragmented configuration of the quartz can be identified via microscopic analysis. No detailed information on these microscopic features were mentioned in his original publication. *Brecciated quartz* in particular became a common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material that functionally behaves similarly to quartzite [e.g. @holm91: 24] (@fig-artefacts). As the material frequently has been characterized on the basis of macroscopic characteristics, despite the fundamental features necessary for characterization being microscopic, there is value in including this material as a test group in the study.

@@

Significant peaks were identified at 1 400 nm, 1 900 nm, 2 200 nm, and 2 350 nm. These features are all the result of overtones and vibration combinations of OH groups, with the two first peaks at 1 400 nm and 1 900 nm being indicative of molecular water  (@fig-nir-peak-summary) [@hunt77: 508-510; @sciuto_etal19]. OH groups can produce a response in a number of different fields in the same material, with metal-OH combinations occurring in the 2 000 nm - 2 500 nm region. While there are exceptions, a more intense 2 200 nm band has typically been related to aluminium hydroxide (AlOH), and a more intense 2 350 nm band to magnesium hydroxide (MgOH) [@hunt77: 509; @clark_etal90: 12664]. In the current dataset the 2 200 nm band is typically the more intense band.

The 2 200 nm and 2 350 nm absorption bands can be found in mica, such as muscovite and biotite, which can be found in quartzite [@clark_etal90: 12; @hunt77: 510; @ramanaidou_etal15: 207]. Although muscovite and biotite share some band positions, the altered Al content as a response to metamorphism has been documented to shift some positions and increase others, such as the 2 250 nm band for biotite [@duke94]. As can be seen in the average spectra in @fig-nir-spectra-summary, the dark material exhibits an additional peak at ~2 250 nm, which may indicate the presence of biotite and a different metamorphic grade.

Two smaller features were recorded, albeit vague and difficult to make out in the non-processed spectra, at 1 065 nm and 1 300 nm (@fig-nir-spectra-summary). These could be caused by ferrous iron (Fe^2+^) [@hunt77: 503-504], which has been demonstrated to produce a response in the NIR in a number of minerals. In studies of olivine the absorption region has been determined to be broad and centred around 1 000 nm, but with three features at 900 nm, 1 100 nm and 1 300 nm [@trang_etal13]. Alternatively, weak absorption features (OH) have also been noted in the 950 nm and 1 150 nm bands, particularly among clays. However, these overlap with the Fe^2+^ bands in, for example, biotite [@adams75: 104-106]. It is possible that the features at 1 065 nm and 1 300 nm can be linked to the presence of ferrous iron, but it is difficult to say what the contributing mineral is at this stage.

@@

In addition to the more intense features described above, a number of weaker peaks can also be seen in the spectra. Some of these could be additional quartz bands [@zhong_etal21], such as a weak peak at 1160 cm^-1^ that is somewhat obscured by the D band. that is obscured by the D band. Two peaks can also be found in the 700 - 800 cm^-1^ band range. The 700 cm^-1^ band has been identified as one of the aforementioned quartz bands. Studies of clay minerals [@kloprogge17], and iron ore minerals [@ramanaidou_etal15], have also generated a response related to Si, Al and OH vibrations within the 700 - 800 cm^-1^ region.

@@

The XRF spectrometer is not capable of detecting all the common elements, and the software therefore calculates a balance (BAL) that represents the remaining unquantifiable elements in a sample. The elements represented in this group are all those with an atomic number lower than ~12~Mg [@thermo18]. The analysis shows that ~14~Si makes up the main bulk of the material, with ~13~Al recorded up to ~ 5% for some samples (@fig-xrf-violin-plot). The majority of the readings resulted in ~14~Si content of less than 50%, as well as a BAL between 40 – 70%, meaning that a significant portion of the material remains unknown. Whilst this could be due to the limitations of the instrumentation, it could also be a result of the sampling. As quartzite is a metamorphic rock where the distribution of quartz is not necessarily uniform in the material, the sample point can have a considerable influence on the measurement. Out of a total of `r length(tot_uniq_samples)` samples `r si_90` measured a ~14~Si content higher than 90%. Certain elements recorded a high number of values below the LOD (“missing values”), including ~12~Mg, vanadium (~23~V), manganese (~25~Mn), yttrium (~39~Y) and zinc (~30~Zn). Although in the case of ~12~Mg and ~25~Mn the recorded values reached up to 0.3% and 3% respectively.

Initial analysis used clr transformed data, but encountered issues related to the structure of the ~14~Si values. Instead log10-transformation was used, producing a similar model, but one that managed the loading for ~14~Si better. The PCA was run twice, firstly with the > 90% ~14~Si samples included and secondly excluded. The initial PCA (model A), with the full dataset included, resulted in 3 PCs with an eigenvalue above 1, and a cumulative explained variance of `r xrf_var`%. K-means cluster analysis was performed on the PCA data in order identify potential groups with similar elemental composition. Using the elbow graph method 3 clusters were deemed suitable for the analysis.

@@

As can be seen in the score plot (@fig-xrf-pca :A), the three K-means clusters remain separated. Cluster 1 is made up of those samples that feature a high ~14~Si content (> 90%). There is no real pattern to this cluster aside from higher ~14~Si values with lower iron (~26~Fe) and zirconium (~40~Zr) values. Loadings for elements expected to follow changes in 14Si content also end up near cluster 1 (e.g. Al, calcium (Ca)).

Cluster 2 is distinguished by a higher titanium (~22~Ti), ~26~Fe and ~40~Zr content. It mainly features samples with a dark hue, with the few samples from the light group being grey in colour and arguably somewhere in between a “dark” or “light” hue. ~13~Al and potassium (~19~K) is similarly higher in cluster 2 than cluster 3, which is in turn marked by its, on average, lower elemental content than the other two clusters. As ~13~Al and ~19~K are indicative of a sedimentary depositional context that has been subject to metamorphism [@hazen_morrison2022], and the cluster with higher values may thus indicate samples with a different grade of metamorphism. Cluster 3 features a more limited range of values between the samples and includes all the *brecciated quartz*.

In the second PCA model (model B) four PCs feature an eigenvalue above 1 with a cumulative explained variance of `r xrf_var_low`%. Three clusters were deemed suitable for the K-means analysis (@fig-xrf-kmeans-boxplot). The range of values is more limited in cluster 3, with overall lower values than compared to the other clusters (@fig-xrf-pca :B). The *brecciated quartz* samples all remain in cluster 3, along with mainly light-to-colourless samples. Cluster 2 features the main bulk of darker samples with higher  ~22~Ti, ~26~Fe and ~40~Zr values. Cluster 1 likewise feature a large portion of the darker materials, with comparable ~22~Ti, ~26~Fe and ~40~Zr values. It also has higher values of ~12~Mg, ~13~Al, ~16~S, ~19~K, and ~20~Ca.

~38~Sr was detected at trace amounts. It is a lithophile element that can substitute for 20Ca in sulphates, feldspars, mica and clays [@ahijado_etal05;@banner95], which might be one explanation for its presence in cluster 1 in model B. Nearly all samples with ~38~Sr are located in cluster 2 in model A.

@@

![XRF PCA score plots: a) Score plot featuring the full dataset, cluster 1 includes all samples with Si > 90%, while the remaining samples seem to cluster based on Fe and Zr content. This results in most of the darker materials clustering towards the bottom of the plot. b) Score plot with > 90% Si samples excluded. Similar to plot A there is a separation based on Fe and Zr values, with darker material towards the bottom of the plot. Light-to-colourless samples remain largely together in cluster 3. Cluster 1 stands out by featuring slightly lower Si, Fe, Ti, and Zr values, but higher values in Al, Ca, K, and Mg](../figures/009-xrf-pca-with-photos_v1.jpg){#fig-xrf-pca width=100%, height=100%}

@@

```{r}
#| label: fig-xrf-kmeans-boxplot
#| cache: true
#| fig-width: 7.87
#| fig-asp: 1
#| out-width: "100%"
#| fig-cap: "Boxplot of XRF values (%) in PCA model B, grouped by K-means cluster. Cluster 1 is made up of samples with lower values of Si, but higher values for those elements that are typically associated with Si (e.g. Al, Ca). Cluster 2 includes the samples with higher Fe/Ti/Zr content, whilst cluster 3 is mainly distinguished by a higher Si content and low Fe/Ti/Zr content"

@@

The K-means cluster analysis of model B is visualised in the NIRS PCA score plot, which further emphasizes the variation within the general colour groups (@fig-nir-xrf-kmeans). Cluster 2, with higher ~26~Fe and ~40~Zr values, mainly coincides with the darker material to the right side of the plot. This was expected as the trend along PC 1 in the NIR PCA mainly separates the material according to absorbance. Cluster 3 in turn coincides with the lighter material in the opposite side of the model. Cluster 1, however, is fairly evenly distributed within the model along PC 1 and PC 2.

@@

The results presented here have increased the state of knowledge of the chemical and structural variation that exists under the umbrella of what archaeologists refer to as quartzite within the Västerbotten area. These results are applicable to neighbouring regions and beyond, and should help in furthering research into prehistoric tools in Scandinavia and other regions. The material exhibits considerable variation in terms of texture and colour, and the results suggest that the initial classification, based on the archaeological documentation, may have included mistakes. These mistakes, whilst of little consequence for site interpretation individually, may introduce considerable noise, or “messiness”, when aggregating data for large scale spatio-temporal analyses [@gattiglia15: 2]. The results also highlight some of the implications that the choice of spectroscopic analysis method may have for the identification of features in a material. In particular, they demonstrate that reliance on any individual method may introduce limitations which could make material classification difficult or misleading.

@@

Whilst NIRS and Raman may be somewhat limited in the characterization of quartz and quartzite they have considerable potential when combined with the XRF. The complementary information about the structural elements in the material can help guide interpretation of the compositional data. Ideally these different sources would be fused using methods such as multi-block PCA to better asses key variables contributing to the model [@mishra_etal21].

@@

We would like to thank Västerbottens museum (www.vbm.se) for access to their collections, and Susanne Sundström in particular for her assistance with preparing the artefacts. We would also like to acknowledge cooperation with the division of Biomass Technology and Chemistry at the Swedish University of Agricultural Sciences.

@@

Johan Linderholm: Conceptualization, Methodology, Supervision

>>>>

---
title: "Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden"
author:
  - Mattias Sjölander:
      email: mattias.sjolander@umu.se
      institute: [UmU]
      correspondence: true
  - Johan Linderholm:
      email: johan.linderholm@umu.se
      institute: [UmU]
      correspondence: false
  - Paul Geladi:
      email: paul.geladi@slu.se
      institute: [SLU]
      correspondence: false
  - Philip I. Buckland:
      email: philip.buckland@umu.se
      institute: [UmU]
      correspondence: false
institute:
  - UmU: Environmental Archaeology Lab, Department of Historical, Philosophical and Religious studies, Umeå University, Umeå, Sweden
  - SLU: Department of Forest Biomaterials and Technology, Swedish University of Agricultural Sciences, Umeå, Sweden 
title-block-published: "Last updated"  
date: now
date-format: long
crossref: 
  fig-title: '**Fig.**'
  fig-labels: '**arabic**'
  fig-prefix: Fig.
format:
  pdf:
    documentclass: scrartcl
    fig-pos: "H"
    geometry: margin = 2.54cm 
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 300
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/quaternary-science-reviews.csl" # Insert path for the bib-style
abstract: |
  Northern Fennoscandia is a geologically complex region affected by both glacial and postglacial processes. Quartzite was a key material type utilized by hunter-gatherers in Northern Sweden around the period 4 000 – 2 000 BP, and is thus critical to the understanding of raw material procurement and material flow within the region. However, there is a severe lack of methodological development in the characterization of these materials, and provenance of locally available geological material is complex and fraught with uncertainty. 126 quartz/quartzite points and preforms were sampled from 47 archaeological sites along the upper Ångerman river valley in Västerbotten, Sweden. The material has been analysed non-destructively using three separate portable spectroscopic instrumentations (Near-infrared, Raman, X-Ray Fluorescence). Evaluation of the spectra and exploratory data analysis using Principal Component Analysis demonstrates detectable differences in the material that likely stem from diagenetic/paragenetic origin. The presence of graphite, muscovite and biotite could likewise provide information on the material’s metamorphic grade. In addition to reaffirming the potential of field-based screening instrumentation, these results will benefit future surveys of geological sources in the region. They also indicate potential for the construction of a predictive model that could classify the quartzite based on its chemical characteristic. Such a model would prove useful in future spatial analysis and testing of models of raw material management.
keywords: |
  Hunter-Gatherer; Lithics; Provenance; Quartzite; Vibrational spectroscopy; Scandinavia
---

@@

Petrographic characterization and provenance studies are instrumental for inferring trade and exchange among prehistoric societies [@tykot03]. The presence of lithic material foreign to the local region of an archaeological site, raises important questions on the processes that resulted in its eventual deposition and discovery. The characterization of geological material and identification of geological sources are the first steps in reconstructing a prehistoric community’s strategies for raw material selection and management. Flint and obsidian have been central to studies of trade and exchange in European prehistory, due in part to the well-surveyed geological sources for these materials [@tykot17]. This focus has led to a material bias which has been highlighted by Prieto et al. [-@prieto_etal19: 15]. Recent decades, however, have seen an increase in interest for non-flint lithics and raw material studies [@blomme_etal12; @columbu_etal23; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19; @ramacciotti_etal22; @ramacciotti_etal23; @sherman_etal23].

In Northern Sweden, quartz and quartzite dominate the lithic assemblages and have been used in settlement, exchange and mobility studies [@baudou78]. Models have been proposed for raw material sourcing and trade and exchange for the less common slate and flint finds within Northern Sweden [@halen94: 115-122; @lundberg97: 161-171]. There has, however, been little development in the petrographic characterization and provenance study of quartz and quartzite [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10]. A possible reason for this might well be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16]. Northern Sweden is extensively covered by glacial till, and whilst there are identifiable source areas (e.g. small bedrock quarries), there is an extremely large number of other potential sources of raw materials to be found in moraine deposits. The few existing theories on lithic material sourcing within the North Swedish region describe strategies for direct procurement from the mountain zone [@forsberg85; -@forsberg12]. This does not, however, account for the possibility that the material was sourced locally around the site, as part of an embedded adaptation strategy [@binford79; @rankama_etal06: 249-250]. A new approach to provenance studies is therefore required in this region.

@@

![Geographical distribution of the sampled sites, with find context, along the river of Ångermanälven, Västerbotten, Sweden. The Late Quaternary reincorporation of geological material from the Scandian mountains into the sediments in the forest inlands adds to the complexity of questions of provenance. Geological symbology after (Gee and Stephens 2020). Made with © Sveriges Geologiska Undersökning, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet.](../figures/Fig1.jpg){#fig-distribution-map width=100%, height=100%}

@@

This study includes `r length(quartzite)` quartzite and `r length(bquartz)` *brecciated quartz* samples. The inclusion of *brecciated quartz* in the study enables the separation of quartz from quartzite, based on non-destructive analysis of their chemical and structural characteristics, to be evaluated. The term  *brecciated quartz* (also referred to as *mylonite quartz*) was introduced to archaeologists in Northern Sweden by the geologist Åhman [-@ahman67: 8] in his analysis of the petrographic material in the *Early Norrland* project [@biornstad68; @ahman67]. Åhman describes the material as easily mistaken for an unusually pure quartzite, but where the fragmented configuration of the quartz can be identified via microscopic analysis. No detailed information on these microscopic features were mentioned in his original publication. *Brecciated quartz* in particular became a common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material that functionally behaves similarly to quartzite [e.g. @holm91: 24] (@fig-artefacts). As the material frequently has been characterized on the basis of macroscopic characteristics, despite the fundamental features necessary for characterization being microscopic, there is value in including this material as a test group in the study.

@@

Significant peaks were identified at 1 400 nm, 1 900 nm, 2 200 nm, and 2 350 nm. These features are all the result of overtones and vibration combinations of OH groups, with the two first peaks at 1 400 nm and 1 900 nm being indicative of molecular water  (@fig-nir-peak-summary) [@hunt77: 508-510; @sciuto_etal19]. Hydroxyl (OH) groups can produce a response in a number of different fields in the same material, with metal-OH combinations occurring in the 2 000 nm - 2 500 nm region. While there are exceptions, a more intense 2 200 nm band has typically been related to aluminium hydroxide (AlOH), and a more intense 2 350 nm band to magnesium hydroxide (MgOH) [@hunt77: 509; @clark_etal90: 12664]. In the current dataset the 2 200 nm band is typically the more intense band.

The 2 200 nm and 2 350 nm absorption bands can be found in mica, such as muscovite and biotite, which can be found in quartzite [@clark_etal90: 12; @hunt77: 510; @ramanaidou_etal15: 207]. Although muscovite and biotite share some band positions, the altered aluminium (Al) content as a response to metamorphism has been documented to shift some positions and increase others, such as the 2 250 nm band for biotite [@duke94]. As can be seen in the average spectra in @fig-nir-spectra-summary, the dark material exhibits an additional peak at ~2 250 nm, which may indicate the presence of biotite and a different metamorphic grade.

Two smaller features were recorded, albeit vague and difficult to make out in the non-processed spectra, at 1 065 nm and 1 300 nm (@fig-nir-peak-summary). These could be caused by ferrous iron (Fe^2+^) [@hunt77: 503-504], which has been demonstrated to produce a response in the NIR in a number of minerals. In studies of olivine the absorption region has been determined to be broad and centred around 1 000 nm, but with three features at 900 nm, 1 100 nm and 1 300 nm [@trang_etal13]. Alternatively, weak absorption features (OH) have also been noted in the 950 nm and 1 150 nm bands, particularly among clays. However, these overlap with the Fe^2+^ bands in, for example, biotite [@adams75: 104-106]. It is possible that the features at 1 065 nm and 1 300 nm can be linked to the presence of ferrous iron, but it is difficult to say what the contributing mineral is at this stage.

@@

In addition to the more intense features described above, a number of weaker peaks can also be seen in the spectra. Some of these could be additional quartz bands [@zhong_etal21], such as a weak peak at 1160 cm^-1^ that is somewhat obscured by the D band. that is obscured by the D band. Two peaks can also be found in the 700 - 800 cm^-1^ band range. The 700 cm^-1^ band has been identified as one of the aforementioned quartz bands. Studies of clay minerals [@kloprogge17], and iron ore minerals [@ramanaidou_etal15], have also generated a response related to silicon (Si), aluminium (Al) and hydroxyl (OH) vibrations within the 700 - 800 cm^-1^ region.

@@

The XRF spectrometer is not capable of detecting all the common elements, and the software therefore calculates a balance (BAL) that represents the remaining unquantifiable elements in a sample. The elements represented in this group are all those with an atomic number lower than magnesium (~12~Mg) [@thermo18]. The analysis shows that silicon (~14~Si) makes up the main bulk of the material, with aluminium (~13~Al) recorded up to ~ 5% for some samples (@fig-xrf-violin-plot). The majority of the readings resulted in silicon (~14~Si) content of less than 50%, as well as a BAL between 40 – 70%, meaning that a significant portion of the material remains unknown. Whilst this could be due to the limitations of the instrumentation, it could also be a result of the sampling. As quartzite is a metamorphic rock where the distribution of quartz is not necessarily uniform in the material, the sample point can have a considerable influence on the measurement. Out of a total of `r length(tot_uniq_samples)` samples `r si_90` measured a silicon (~14~Si) content higher than 90%. Certain elements recorded a high number of values below the LOD (“missing values”), including magnesium (~12~Mg), vanadium (~23~V), manganese (~25~Mn), yttrium (~39~Y) and zinc (~30~Zn). Although in the case of magnesium (~12~Mg) and manganese (~25~Mn) the recorded values reached up to 0.3% and 3% respectively.

Initial analysis used clr transformed data, but encountered issues related to the structure of the silicon (~14~Si) values. Instead log10-transformation was used, producing a similar model, but one that managed the loading for silicon (~14~Si) better. The PCA was run twice, firstly with the > 90% silicon (~14~Si) samples included and secondly excluded. The initial PCA (model A), with the full dataset included, resulted in 3 PCs with an eigenvalue above 1, and a cumulative explained variance of `r xrf_var`%. K-means cluster analysis was performed on the PCA data in order identify potential groups with similar elemental composition. Using the elbow graph method 3 clusters were deemed suitable for the analysis.

@@

As can be seen in the score plot (@fig-xrf-pca :A), the three K-means clusters remain separated. Cluster 1 is made up of those samples that feature a high silicon (~14~Si) content (> 90%). There is no real pattern to this cluster aside from higher silicon (~14~Si) values with lower iron (~26~Fe) and zirconium (~40~Zr) values. Loadings for elements expected to follow changes in silicon( ~14~Si) content also end up near cluster 1 (e.g. aluminium (~13~Al), calcium (~20~Ca)).

Cluster 2 is distinguished by a higher titanium (~22~Ti), iron (~26~Fe) and zirconium (~40~Zr) content. It mainly features samples with a dark hue, with the few samples from the light group being grey in colour and arguably somewhere in between a “dark” or “light” hue. Aluminium (~13~Al) and potassium (~19~K) is similarly higher in cluster 2 than cluster 3, which is in turn marked by its, on average, lower elemental content than the other two clusters. As aluminium (~13~Al) and potassium (~19~K) are indicative of a sedimentary depositional context that has been subject to metamorphism [@hazen_morrison2022], and the cluster with higher values may thus indicate samples with a different grade of metamorphism. Cluster 3 features a more limited range of values between the samples and includes all the *brecciated quartz*.

In the second PCA model (model B) four PCs feature an eigenvalue above 1 with a cumulative explained variance of `r xrf_var_low`%. Three clusters were deemed suitable for the K-means analysis (@fig-xrf-kmeans-boxplot). The range of values is more limited in cluster 3, with overall lower values than compared to the other clusters (@fig-xrf-pca :B). The *brecciated quartz* samples all remain in cluster 3, along with mainly light-to-colourless samples. Cluster 2 features the main bulk of darker samples with higher titanium (~22~Ti), iron (~26~Fe) and zirconium (~40~Zr) values. Cluster 1 likewise feature a large portion of the darker materials, with comparable titanium (~22~Ti), iron (~26~Fe) and zirconium (~40~Zr) values. It also has higher values of magnesium (~12~Mg), aluminium (~13~Al), sulphur (~16~S), potassium (~19~K), and calcium (~20~Ca).

Strontium (~38~Sr) was detected at trace amounts. It is a lithophile element that can substitute for calcium (~20~Ca) in sulphates, feldspars, mica and clays [@ahijado_etal05;@banner95], which might be one explanation for its presence in cluster 1 in model B. Nearly all samples with strontium (~38~Sr) are located in cluster 2 in model A.

@@

![XRF PCA score plots: a) Score plot featuring the full dataset, cluster 1 includes all samples with Si > 90%, while the remaining samples seem to cluster based on Fe and Zr content. This results in most of the darker materials clustering towards the bottom of the plot. b) Score plot with > 90% Si samples excluded. Similar to plot A there is a separation based on ~26~Fe and ~40~Zr values, with darker material towards the bottom of the plot. Light-to-colourless samples remain largely together in cluster 3. Cluster 1 stands out by featuring slightly lower ~14~Si, ~26~Fe, ~22~Ti, and ~40~Zr values, but higher values in ~13~Al, ~20~Ca, ~19~K, and ~12~Mg](../figures/009-xrf-pca-with-photos_v1.jpg){#fig-xrf-pca width=100%, height=100%}

@@

```{r}
#| label: fig-xrf-kmeans-boxplot
#| cache: true
#| fig-width: 7.87
#| fig-asp: 1
#| out-width: "100%"
#| fig-cap: "Boxplot of XRF values (%) in PCA model B, grouped by K-means cluster. Cluster 1 is made up of samples with lower values of ~14~Si, but higher values for those elements that are typically associated with Si (e.g. ~13~Al, ~20~Ca). Cluster 2 includes the samples with higher ~26~Fe/~22~Ti/~40~Zr content, whilst cluster 3 is mainly distinguished by a higher Si content and low ~26~Fe/~22~Ti/~40~Zr content"

@@

The K-means cluster analysis of model B is visualised in the NIRS PCA score plot, which further emphasizes the variation within the general colour groups (@fig-nir-xrf-kmeans). Cluster 2, with higher iron (~26~Fe) and zirconium (~40~Zr) values, mainly coincides with the darker material to the right side of the plot. This was expected as the trend along PC 1 in the NIR PCA mainly separates the material according to absorbance. Cluster 3 in turn coincides with the lighter material in the opposite side of the model. Cluster 1, however, is fairly evenly distributed within the model along PC 1 and PC 2.

@@

The results presented here have increased the state of knowledge of the chemical and structural variation that exists under the umbrella of what archaeologists refer to as quartzite within the Västerbotten area. This non-destructive exploratory approach is particularly useful in regions where there is a lack of data and an element of uncertainty (see Jebb et al., [-@jebb_etal17] for discussion of Exploratory Data Analysis), making it possible to identify variables and trends of interest to follow up with more deductive methods. The same approach can be applied in most regions where there is a lack of information on which to build a robust provenance study. The main limitation is the material in question and how it responds to the spectroscopic instrumentation. The results are applicable to neighbouring regions and beyond, and should help in furthering research into prehistoric tools in Scandinavia and other areas. The material exhibits considerable variation in terms of texture and colour, and the results suggest that the initial classification, based on the archaeological documentation, may have included mistakes. These mistakes, whilst of little consequence for site interpretation individually, may introduce considerable noise, or “messiness”, when aggregating data for large scale spatio-temporal analyses [@gattiglia15: 2]. The results also highlight some of the implications that the choice of spectroscopic analysis method may have for the identification of features in a material. In particular, they demonstrate that reliance on any individual method may introduce limitations which could make material classification difficult or misleading.

@@

Whilst NIRS and Raman may be somewhat limited in the characterization of quartz and quartzite they have considerable potential when combined with the XRF. The complementary information about the structural elements in the material can help guide interpretation of the compositional data. Ideally these different sources would be fused using methods such as multi-block PCA to better asses key variables contributing to the model [@mishra_etal21]. Such analysis is, however, outside the scope of this paper and a potential direction for future work.

@@

We would like to thank Västerbottens museum (www.vbm.se) for access to their collections, and Susanne Sundström in particular for her assistance with preparing the artefacts. We would also like to acknowledge cooperation with the division of Biomass Technology and Chemistry at the Swedish University of Agricultural Sciences. Finally, we would like to thank the detailed comments and suggestions for improvement provided by two anonymous reviewers.

@@

Johan Linderholm: Conceptualization, Methodology, Supervision, Writing - Review & Editing
